---
title: "Getting and Cleaning Data Course Project Codebook"
output:
  html_document:
    keep_md: yes
---

## Project Description
The project was to analyse the data captured via an accelerometer and gyrascope within a smartphone and use this to create a tidy data set consisting of mean and standard deviation readings for 6 activities undertaken by each of 30 volunteers.

##Study design and data processing

###Collection of the raw data
The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, 3-axial linear acceleration and 3-axial angular velocity were captured at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The original dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. 

###Notes on the original (raw) data 
The raw data consisted of 10,299 observations of 561 variables from 30 volunteers undertaking 6 activities. 

##Creating the tidy datafile

###Guide to create the tidy data file
To create the tidy data file, simply download the script 'run_analysis.R' into the working directory, load it using the command

```{r eval=FALSE}
source("run_analysis.R")
```

and then run with

```{r eval=FALSE}
run_analysis()
```

The relevant files will be downloaded, extracted and a tidy data set, called 'tidy-data.txt' will be produced. 

###Cleaning of the data
The files downloaded are unzipped, grouped together into a single data frame and tidied by adding descriptive elements to the variables and activities that the participants undertook. 

A detailed description of the cleaning process is included in the README.md file. 

##Description of the variables in the tidy-data.txt file
The tidy data contains 180 observations (30 participants x 6 activities) across 66 variables. The dimensions of the dataset are 180 x 68 with the extra two columns being used to identify the 'Subject' (participant) and 'Activity' being measured. 

- Subject is one of the numbers 1:30 corresponding to a particular participant
- Activity is one of six activities 
- Each of the remaining variables is either a mean or standard deviation of the data for that participant undertaking a particular activity. The values, x, all fall in the range -1 < x < 1 due to the norm process (which normalises results based on the number of observations). 

The individual variable names are:

```{r echo=FALSE}
tidy_data <- read.table("tidy-data.txt", header=TRUE)
names(tidy_data)
```

###Variable 1 - Subject
This is the ID number of the subject under observation. It is an integer field ranging from 1 to 30 for the 30 participants. 

###Variable 2 - Activity
These are the 6 activities that each participant undertook. This is a factor variable with levels:

* WALKING
* WALKING_UPSTAIRS
* WALKING_DOWNSTAIRS
* SITTING
* STANDING
* LAYING

###Variables 3:68 - Measurements
These are 66 different means and standard deviations of the various variables under observation. They are categorised with descriptive names including x, y and z (whether they pertain to a particular axis movement), and mean/std for whether they are a mean or standard deviation reading respectively. All of these measurements are numeric and fall within the range (-1,1), each one in units of Hz. 

```{r echo=FALSE}
tidy_data <- read.table("tidy-data.txt", header=TRUE)
str(tidy_data[,3:66])
```

##Sources
The dataset is a summary from the 'Human Activity Recognition Using Smartphones Dataset Version 1.0', obtained from 

[http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)

Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto.
Smartlab - Non Linear Complex Systems Laboratory
DITEN - Università degli Studi di Genova.
Via Opera Pia 11A, I-16145, Genoa, Italy.
activityrecognition@smartlab.ws
www.smartlab.ws

For more information about this dataset contact: activityrecognition@smartlab.ws

License:
========
Use of this dataset in publications must be acknowledged by referencing the following publication [1] 

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.

##Annex
Code used in this codebook is as follows.

To obtain the names of the variables in the tidy-data.txt file:
```{r eval=FALSE}
tidy_data <- read.table("tidy-data.txt", header=TRUE)
names(tidy_data)
```

To list the summary of all variables in the tidy-data.txt file:
```{r eval=FALSE}
tidy_data <- read.table("tidy-data.txt", header=TRUE)
str(tidy_data[,3:66])
```
